### 2017-11-02 submission/manuscript.Rmd
add second discussion paragraph on driver-passenger

- Flynn review and how data supports an idea of early specific localized community changes that becomes global in CRC

>>>>

The driver-passenger model of the microbial role in CRC, as summarized by Flynn [@flynn_metabolic_2016], can be supported with this data for carcinoma but not necessarily for adenoma. The drasitically increased relative risk of disease when considering the CRC associated genera is highly supportive of this type of process. In a driver-passenger scenario it is possible that simply having the driver present or only identifying the passenger is a good enough proxy that the event is occuring. This would account for the observation that there is no synergistic incrase in relative risk when accounting for either the total number or increasing abundance of these genera. The initial establishment of the driver within the system is also dependent on the community that is present and this could explain why adding the community context to our models increases the prediction. 
Although the driver-passenger model fits well with the transtion of adenoma to carcinoma for our carcinoma observations it does not fit well with our observed data on control to adenoma.

### 2017-11-02 submission/manuscript.tex
add second discussion paragraph on driver-passenger

\begin{itemize}
\item
  Flynn review and how data supports an idea of early specific localized
  community changes that becomes global in CRC
\item
  Builiding on this idea moving forward how CRC associated genera
  interact with the overall community rather than in isolation will be
  important since they do not increase the RR of polyps
\item
  suggesting localized changes to community may be more important
\item
  In closing to get at these more difficult questions the Power
  calculations and effect size in future studies need to be accounted
  for especially for adenoma where the predicted effect sizes are going
  to be small
\end{itemize}

@@

\textbf{\emph{Obtaining Data Sets:}} Studies used for this meta-analysis
were identified through the review articles written by Keku, et al. and
Vogtmann, et al. {[}1,2{]}. All studies were included that used tissue
or feces as their sample source for 16S rRNA gene sequencing analysis.
Studies using either 454 or Illumina sequencing technology were
included. Only data sets that had the raw sequences available for
analysis were included. Some studies did not have publically available
raw sequences or did not have meta data in which the authors were able
to share. After this filtering step the following studies remained: Ahn
{[}3{]}, Baxter {[}4{]}, Brim {[}5{]}, Burns {[}6{]}, Chen {[}7{]},
Dejea {[}8{]}, Flemer {[}9{]}, Geng {[}10{]}, Hale {[}11{]}, Kostic
{[}12{]}, Lu {[}13{]}, Sanapareddy {[}14{]}, Wang {[}15{]}, Weir
{[}16{]}, and Zeller {[}17{]}. The Zackular {[}18{]} study was not
included becasue the 90 individuals analyzed within the study are
contained within the larger Baxter study. The Kostic study was not used
since after sequence processing all the case samples did not have more
than 100 sequences remaining. This left a total of 13 studies in which
complete analysis could be completed.

@@

\textbf{\emph{Sequence Processing:}} For the majority of studies raw
sequences were downloaded from the SRA
(\url{ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/})
and metadata was obtained from the following website:
\url{http://www.ncbi.nlm.nih.gov/Traces/study/} by searching the
respective accession number of the study. Of the studies that did not
have sequences and meta data on the SRA one study had the data stored on
DBGap {[}3{]} and four studies the data was obtained directly from the
authors {[}9,11,14,16{]}. Each study was processed using the mothur
(v1.39.3) software program {[}19{]}. Where possible quality filtering
utilized the default methods used in mothur for either 454 or Illumina
based sequencing. If it was not possible to use these defaults the
author stated quality cut-offs were used instead. Chimeras were
identifed and removed using the VSEARCH {[}20{]} program and \emph{de
novo} OTU clustering at 97\% similarity using the OptiClust algorithm
{[}21{]} was utilized.

@@

\hypertarget{refs}{}
\hypertarget{ref-keku_gastrointestinal_2015}{}
1. Keku TO, Dulal S, Deveaux A, Jovov B, Han X. The gastrointestinal
microbiota and colorectal cancer. American Journal of Physiology -
Gastrointestinal and Liver Physiology {[}Internet{]}. 2015 {[}cited 2017
Oct 30{]};308:G351--63. Available from:
\url{http://ajpgi.physiology.org/lookup/doi/10.1152/ajpgi.00360.2012}

\hypertarget{ref-vogtmann_epidemiologic_2016}{}
2. Vogtmann E, Goedert JJ. Epidemiologic studies of the human microbiome
and cancer. British Journal of Cancer {[}Internet{]}. 2016 {[}cited 2017
Oct 30{]};114:237--42. Available from:
\url{http://www.nature.com/doifinder/10.1038/bjc.2015.465}

\hypertarget{ref-ahn_human_2013}{}
3. Ahn J, Sinha R, Pei Z, Dominianni C, Wu J, Shi J, et al. Human gut
microbiome and risk for colorectal cancer. Journal of the National
Cancer Institute. 2013;105:1907--11.

\hypertarget{ref-baxter_microbiota-based_2016}{}
4. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
improves the sensitivity of fecal immunochemical test for detecting
colonic lesions. Genome Medicine. 2016;8:37.

\hypertarget{ref-brim_microbiome_2013}{}
5. Brim H, Yooseph S, Zoetendal EG, Lee E, Torralbo M, Laiyemo AO, et
al. Microbiome analysis of stool samples from African Americans with
colon polyps. PloS One. 2013;8:e81352.

\hypertarget{ref-burns_virulence_2015}{}
6. Burns MB, Lynch J, Starr TK, Knights D, Blekhman R. Virulence genes
are a signature of the microbiome in the colorectal tumor
microenvironment. Genome Medicine. 2015;7:55.

\hypertarget{ref-chen_human_2012}{}
7. Chen W, Liu F, Ling Z, Tong X, Xiang C. Human intestinal lumen and
mucosa-associated microbiota in patients with colorectal cancer. PloS
One. 2012;7:e39743.

\hypertarget{ref-dejea_microbiota_2014}{}
8. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
Rossetti BJ, et al. Microbiota organization is a distinct feature of
proximal colorectal cancers. Proceedings of the National Academy of
Sciences of the United States of America. 2014;111:18321--6.

\hypertarget{ref-flemer_tumour-associated_2017}{}
9. Flemer B, Lynch DB, Brown JMR, Jeffery IB, Ryan FJ, Claesson MJ, et
al. Tumour-associated and non-tumour-associated microbiota in colorectal
cancer. Gut. 2017;66:633--43.

\hypertarget{ref-geng_diversified_2013}{}
10. Geng J, Fan H, Tang X, Zhai H, Zhang Z. Diversified pattern of the
human colorectal cancer microbiome. Gut Pathogens. 2013;5:2.

\hypertarget{ref-hale_shifts_2017}{}
11. Hale VL, Chen J, Johnson S, Harrington SC, Yab TC, Smyrk TC, et al.
Shifts in the Fecal Microbiota Associated with Adenomatous Polyps.
Cancer Epidemiology, Biomarkers \& Prevention: A Publication of the
American Association for Cancer Research, Cosponsored by the American
Society of Preventive Oncology. 2017;26:85--94.

\hypertarget{ref-kostic_genomic_2012}{}
12. Kostic AD, Gevers D, Pedamallu CS, Michaud M, Duke F, Earl AM, et
al. Genomic analysis identifies association of Fusobacterium with
colorectal carcinoma. Genome Research. 2012;22:292--8.

\hypertarget{ref-lu_mucosal_2016}{}
13. Lu Y, Chen J, Zheng J, Hu G, Wang J, Huang C, et al. Mucosal
adherent bacterial dysbiosis in patients with colorectal adenomas.
Scientific Reports. 2016;6:26337.

\hypertarget{ref-sanapareddy_increased_2012}{}
14. Sanapareddy N, Legge RM, Jovov B, McCoy A, Burcal L, Araujo-Perez F,
et al. Increased rectal microbial richness is associated with the
presence of colorectal adenomas in humans. The ISME journal.
2012;6:1858--68.

\hypertarget{ref-wang_structural_2012}{}
15. Wang T, Cai G, Qiu Y, Fei N, Zhang M, Pang X, et al. Structural
segregation of gut microbiota between colorectal cancer patients and
healthy volunteers. The ISME journal. 2012;6:320--9.

\hypertarget{ref-weir_stool_2013}{}
16. Weir TL, Manter DK, Sheflin AM, Barnett BA, Heuberger AL, Ryan EP.
Stool microbiome and metabolome differences between colorectal cancer
patients and healthy adults. PloS One. 2013;8:e70803.

\hypertarget{ref-zeller_potential_2014}{}
17. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, et al.
Potential of fecal microbiota for early-stage detection of colorectal
cancer. Molecular Systems Biology. 2014;10:766.

\hypertarget{ref-zackular_human_2014}{}
18. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. The human gut
microbiome as a screening tool for colorectal cancer. Cancer Prevention
Research (Philadelphia, Pa.). 2014;7:1112--21.

\hypertarget{ref-schloss_introducing_2009}{}
19. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
EB, et al. Introducing mothur: Open-Source, Platform-Independent,
Community-Supported Software for Describing and Comparing Microbial
Communities. Appl.Environ.Microbiol. {[}Internet{]}. 2009 {[}cited 12AD
Jan 1{]};75:7537--41. Available from:
\url{http://aem.asm.org/cgi/content/abstract/75/23/7537}

\hypertarget{ref-rognes_vsearch_2016}{}
20. Rognes T, Flouri T, Nichols B, Quince C, Mahé F. VSEARCH: A
versatile open source tool for metagenomics. PeerJ. 2016;4:e2584.

>>>>

The driver-passenger model of the microbial role in CRC, as summarized
by Flynn {[}1{]}, can be supported with this data for carcinoma but not
necessarily for adenoma. The drasitically increased relative risk of
disease when considering the CRC associated genera is highly supportive
of this type of process. In a driver-passenger scenario it is possible
that simply having the driver present or only identifying the passenger
is a good enough proxy that the event is occuring. This would account
for the observation that there is no synergistic incrase in relative
risk when accounting for either the total number or increasing abundance
of these genera. The initial establishment of the driver within the
system is also dependent on the community that is present and this could
explain why adding the community context to our models increases the
prediction. Although the driver-passenger model fits well with the
transtion of adenoma to carcinoma for our carcinoma observations it does
not fit well with our observed data on control to adenoma.

\begin{itemize}
\item
  Builiding on this idea moving forward how CRC associated genera
  interact with the overall community rather than in isolation will be
  important since they do not increase the RR of polyps
\item
  suggesting localized changes to community may be more important
\item
  In closing to get at these more difficult questions the Power
  calculations and effect size in future studies need to be accounted
  for especially for adenoma where the predicted effect sizes are going
  to be small
\end{itemize}

@@

\textbf{\emph{Obtaining Data Sets:}} Studies used for this meta-analysis
were identified through the review articles written by Keku, et al. and
Vogtmann, et al. {[}2,3{]}. All studies were included that used tissue
or feces as their sample source for 16S rRNA gene sequencing analysis.
Studies using either 454 or Illumina sequencing technology were
included. Only data sets that had the raw sequences available for
analysis were included. Some studies did not have publically available
raw sequences or did not have meta data in which the authors were able
to share. After this filtering step the following studies remained: Ahn
{[}4{]}, Baxter {[}5{]}, Brim {[}6{]}, Burns {[}7{]}, Chen {[}8{]},
Dejea {[}9{]}, Flemer {[}10{]}, Geng {[}11{]}, Hale {[}12{]}, Kostic
{[}13{]}, Lu {[}14{]}, Sanapareddy {[}15{]}, Wang {[}16{]}, Weir
{[}17{]}, and Zeller {[}18{]}. The Zackular {[}19{]} study was not
included becasue the 90 individuals analyzed within the study are
contained within the larger Baxter study. The Kostic study was not used
since after sequence processing all the case samples did not have more
than 100 sequences remaining. This left a total of 13 studies in which
complete analysis could be completed.

@@

\textbf{\emph{Sequence Processing:}} For the majority of studies raw
sequences were downloaded from the SRA
(\url{ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/})
and metadata was obtained from the following website:
\url{http://www.ncbi.nlm.nih.gov/Traces/study/} by searching the
respective accession number of the study. Of the studies that did not
have sequences and meta data on the SRA one study had the data stored on
DBGap {[}4{]} and four studies the data was obtained directly from the
authors {[}10,12,15,17{]}. Each study was processed using the mothur
(v1.39.3) software program {[}20{]}. Where possible quality filtering
utilized the default methods used in mothur for either 454 or Illumina
based sequencing. If it was not possible to use these defaults the
author stated quality cut-offs were used instead. Chimeras were
identifed and removed using the VSEARCH {[}{\textbf{???}}{]} program and
\emph{de novo} OTU clustering at 97\% similarity using the OptiClust
algorithm {[}21{]} was utilized.

@@

\hypertarget{refs}{}
\hypertarget{ref-flynn_metabolic_2016}{}
1. Flynn KJ, Baxter NT, Schloss PD. Metabolic and Community Synergy of
Oral Bacteria in Colorectal Cancer. mSphere. 2016;1.

\hypertarget{ref-keku_gastrointestinal_2015}{}
2. Keku TO, Dulal S, Deveaux A, Jovov B, Han X. The gastrointestinal
microbiota and colorectal cancer. American Journal of Physiology -
Gastrointestinal and Liver Physiology {[}Internet{]}. 2015 {[}cited 2017
Oct 30{]};308:G351--63. Available from:
\url{http://ajpgi.physiology.org/lookup/doi/10.1152/ajpgi.00360.2012}

\hypertarget{ref-vogtmann_epidemiologic_2016}{}
3. Vogtmann E, Goedert JJ. Epidemiologic studies of the human microbiome
and cancer. British Journal of Cancer {[}Internet{]}. 2016 {[}cited 2017
Oct 30{]};114:237--42. Available from:
\url{http://www.nature.com/doifinder/10.1038/bjc.2015.465}

\hypertarget{ref-ahn_human_2013}{}
4. Ahn J, Sinha R, Pei Z, Dominianni C, Wu J, Shi J, et al. Human gut
microbiome and risk for colorectal cancer. Journal of the National
Cancer Institute. 2013;105:1907--11.

\hypertarget{ref-baxter_microbiota-based_2016}{}
5. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
improves the sensitivity of fecal immunochemical test for detecting
colonic lesions. Genome Medicine. 2016;8:37.

\hypertarget{ref-brim_microbiome_2013}{}
6. Brim H, Yooseph S, Zoetendal EG, Lee E, Torralbo M, Laiyemo AO, et
al. Microbiome analysis of stool samples from African Americans with
colon polyps. PloS One. 2013;8:e81352.

\hypertarget{ref-burns_virulence_2015}{}
7. Burns MB, Lynch J, Starr TK, Knights D, Blekhman R. Virulence genes
are a signature of the microbiome in the colorectal tumor
microenvironment. Genome Medicine. 2015;7:55.

\hypertarget{ref-chen_human_2012}{}
8. Chen W, Liu F, Ling Z, Tong X, Xiang C. Human intestinal lumen and
mucosa-associated microbiota in patients with colorectal cancer. PloS
One. 2012;7:e39743.

\hypertarget{ref-dejea_microbiota_2014}{}
9. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
Rossetti BJ, et al. Microbiota organization is a distinct feature of
proximal colorectal cancers. Proceedings of the National Academy of
Sciences of the United States of America. 2014;111:18321--6.

\hypertarget{ref-flemer_tumour-associated_2017}{}
10. Flemer B, Lynch DB, Brown JMR, Jeffery IB, Ryan FJ, Claesson MJ, et
al. Tumour-associated and non-tumour-associated microbiota in colorectal
cancer. Gut. 2017;66:633--43.

\hypertarget{ref-geng_diversified_2013}{}
11. Geng J, Fan H, Tang X, Zhai H, Zhang Z. Diversified pattern of the
human colorectal cancer microbiome. Gut Pathogens. 2013;5:2.

\hypertarget{ref-hale_shifts_2017}{}
12. Hale VL, Chen J, Johnson S, Harrington SC, Yab TC, Smyrk TC, et al.
Shifts in the Fecal Microbiota Associated with Adenomatous Polyps.
Cancer Epidemiology, Biomarkers \& Prevention: A Publication of the
American Association for Cancer Research, Cosponsored by the American
Society of Preventive Oncology. 2017;26:85--94.

\hypertarget{ref-kostic_genomic_2012}{}
13. Kostic AD, Gevers D, Pedamallu CS, Michaud M, Duke F, Earl AM, et
al. Genomic analysis identifies association of Fusobacterium with
colorectal carcinoma. Genome Research. 2012;22:292--8.

\hypertarget{ref-lu_mucosal_2016}{}
14. Lu Y, Chen J, Zheng J, Hu G, Wang J, Huang C, et al. Mucosal
adherent bacterial dysbiosis in patients with colorectal adenomas.
Scientific Reports. 2016;6:26337.

\hypertarget{ref-sanapareddy_increased_2012}{}
15. Sanapareddy N, Legge RM, Jovov B, McCoy A, Burcal L, Araujo-Perez F,
et al. Increased rectal microbial richness is associated with the
presence of colorectal adenomas in humans. The ISME journal.
2012;6:1858--68.

\hypertarget{ref-wang_structural_2012}{}
16. Wang T, Cai G, Qiu Y, Fei N, Zhang M, Pang X, et al. Structural
segregation of gut microbiota between colorectal cancer patients and
healthy volunteers. The ISME journal. 2012;6:320--9.

\hypertarget{ref-weir_stool_2013}{}
17. Weir TL, Manter DK, Sheflin AM, Barnett BA, Heuberger AL, Ryan EP.
Stool microbiome and metabolome differences between colorectal cancer
patients and healthy adults. PloS One. 2013;8:e70803.

\hypertarget{ref-zeller_potential_2014}{}
18. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, et al.
Potential of fecal microbiota for early-stage detection of colorectal
cancer. Molecular Systems Biology. 2014;10:766.

\hypertarget{ref-zackular_human_2014}{}
19. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. The human gut
microbiome as a screening tool for colorectal cancer. Cancer Prevention
Research (Philadelphia, Pa.). 2014;7:1112--21.

\hypertarget{ref-schloss_introducing_2009}{}
20. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
EB, et al. Introducing mothur: Open-Source, Platform-Independent,
Community-Supported Software for Describing and Comparing Microbial
Communities. Appl.Environ.Microbiol. {[}Internet{]}. 2009 {[}cited 12AD
Jan 1{]};75:7537--41. Available from:
\url{http://aem.asm.org/cgi/content/abstract/75/23/7537}
